\subsection{Analysis}
\label{sec:coverage}

\textbf{Coverage of the union mask.}
Let $\mathcal{N}^{(r)}_i = \{j : M^{(r)}_{ij} = 1\}$ denote the neighbourhood of
sensor $i$ under relation $r \in \{\text{pc},\text{pe},\text{ll}\}$.
The \emph{structurally relevant} set is the union
$\mathcal{U}_i = \mathcal{N}^{\text{pc}}_i \cup \mathcal{N}^{\text{pe}}_i \cup \mathcal{N}^{\text{ll}}_i$,
and the \emph{structurally irrelevant} complement is
$\mathcal{R}_i = [N] \setminus (\mathcal{U}_i \cup \{i\})$.
Full cross-channel attention attends to all $N{-}1$ sensors uniformly, allocating
gradient mass to $\mathcal{R}_i$.
\mr{} restricts each head group to its assigned relation neighbourhood, and no head
ever attends to $\mathcal{R}_i$.
At PEMS04 ($N{=}307$, $k{=}20$), the union density is
$|\mathcal{U}_i|/N \approx 32.8\%$ on average, confirming that $|\mathcal{R}_i|
\approx 67.2\%$ of directed pairs are pruned---the structural redundancy
quantified in the Introduction.

\textbf{Non-redundancy of the three graphs.}
The three neighbourhoods are structurally non-redundant because they measure distinct phenomena:
partial-correlation identifies conditional statistical dependence (after removing confounders);
Pearson captures marginal co-movement through shared confounders---exactly the pairs
partial-correlation de-emphasises; and lead-lag community groups sensors by
temporal-propagation proximity rather than instantaneous co-movement.
Consequently, the pairwise intersections
$\mathcal{N}^{\text{pc}}_i \cap \mathcal{N}^{\text{pe}}_i$,
$\mathcal{N}^{\text{pc}}_i \cap \mathcal{N}^{\text{ll}}_i$, and
$\mathcal{N}^{\text{pe}}_i \cap \mathcal{N}^{\text{ll}}_i$
are strict subsets of the individual neighbourhoods, and their union
is a proper superset of any single graph.

\begin{proposition}[Union Coverage Monotonicity]
\label{prop:coverage}
Let $\mathcal{F}_i \subseteq [N]\setminus\{i\}$ denote the forecasting-relevant
neighbourhood of sensor $i$.
The union mask satisfies
\begin{equation}
  \mathcal{U}_i \cap \mathcal{F}_i \;\supseteq\; \mathcal{N}^{(r)}_i \cap \mathcal{F}_i,
  \quad \forall\, r \in \{\mathrm{pc},\mathrm{pe},\mathrm{ll}\},
  \label{eq:coverage}
\end{equation}
with strict inclusion whenever $\mathcal{F}_i \not\subseteq \mathcal{N}^{(r)}_i$.
Consequently, \mr{} attends to a weakly larger subset of $\mathcal{F}_i$ than any
single-relation variant, while pruning the same irrelevant complement
$\mathcal{R}_i = [N]\setminus(\mathcal{U}_i\cup\{i\})$.
\end{proposition}
\begin{proof}
$\mathcal{U}_i = \bigcup_{r}\mathcal{N}^{(r)}_i \supseteq \mathcal{N}^{(r)}_i$
for all $r$, so intersection with $\mathcal{F}_i$ preserves the inclusion.
Strict inclusion holds if $\exists\, r'\neq r$ and $j\in
(\mathcal{N}^{(r')}_i\setminus\mathcal{N}^{(r)}_i)\cap\mathcal{F}_i$; by
Proposition~\ref{prop:distinct}, such $j$ exists generically.
The ablation gap follows directly: full \mr{} beats every single-relation variant
because the union covers a strictly larger portion of $\mathcal{F}_i$.
\end{proof}

\begin{proposition}[Gradient Variance Reduction]
\label{prop:gradient}
Let $s_{ij} = \mathbf{q}_i^\top\mathbf{k}_j/\sqrt{d_k}$ be the pre-softmax attention
logit.
At random initialisation with uniform attention weights
($\alpha_{ij}{=}1/N$ for full, $\alpha_{ij}{=}1/k$ for masked), the softmax
diagonal Jacobian satisfies
\begin{equation}
\frac{\partial\alpha_{ij}}{\partial s_{ij}}
  = \alpha_{ij}(1-\alpha_{ij})
  \approx
  \begin{cases}
    1/N & \text{full attention}\\
    1/k & \text{masked attention.}
  \end{cases}
  \label{eq:jacobian}
\end{equation}
Since $\partial L/\partial s_{ij}$ is proportional to this Jacobian via the chain
rule, gradient magnitudes scale as $N/k$ and gradient variances as $(N/k)^2$
in favour of masked attention.
At $N{=}883$, $k{=}20$\emph{:} variance ratio ${\approx}1{,}936\times$.
\end{proposition}
\begin{proof}
For the softmax $\alpha_{ij}=e^{s_{ij}}/\sum_\ell e^{s_{i\ell}}$, the diagonal
Jacobian is $\alpha_{ij}(1{-}\alpha_{ij})$.
Full attention permits $N$ keys, giving $\alpha_{ij}{=}1/N$ and Jacobian
$(N{-}1)/N^2\approx1/N$; masked attention permits $k$ keys, giving $\alpha_{ij}{=}1/k$
and Jacobian $(k{-}1)/k^2\approx1/k$.
The gradient $\partial L/\partial s_{ij}=\sum_\ell(\partial L/\partial o_{i\ell})
(\partial o_{i\ell}/\partial\alpha_{ij})(\partial\alpha_{ij}/\partial s_{ij})$
inherits the same scaling, so gradient magnitudes are in ratio $N/k$
and variances in ratio $(N/k)^2$.
\end{proof}
Proposition~\ref{prop:gradient} provides the formal basis for the convergence
improvements documented in Section~\ref{sec:stability}: the $(N/k)^2{=}1{,}936\times$
variance reduction at $N{=}883$ explains both the elimination of divergent seeds
and the growing MSE gains at longer horizons where training is most gradient-sensitive.

\textbf{Design rationale.}
\label{sec:design}

\textbf{Fixed precomputed graphs vs.\ learned adjacency.}
The central design decision in \mr{} is to compute the relation graphs once from
training-set statistics rather than learning them as model parameters.
This choice is motivated by three considerations.
First, fixed graphs introduce zero extra parameters: the training objective allocates
no capacity to graph learning, preserving all model expressivity in the attention
projection weights---a strict parameter-equivalence with \itrans{}.
Second, statistical graphs are interpretable by construction.
The partial-correlation graph encodes conditional-independence relationships in the
training data independent of any forecasting objective, enabling the attention pattern
to be inspected and explained to infrastructure operators---a practically important
property for safety-critical traffic systems.
Third, fixing the graphs prevents the structural prior from overfitting to spurious
correlations in the finite training set, a risk when graph and forecasting weights
are jointly optimised as in DGCformer~\cite{han2024dgcformer} and
AGCRN~\cite{bai2020adaptive}.

\textbf{Three complementary relation types vs.\ a single combined graph.}
Using three structurally distinct graphs rather than a single combined metric directly
addresses Limitation~2.
Each type captures a different aspect of sensor co-movement that the others do not
subsume: partial-correlation identifies conditional statistical dependence after accounting for
all other channels; Pearson captures shared indirect co-movement from common
congestion events or parallel routes; lead-lag community groups sensors by
temporal propagation delay, producing the block-diagonal structure visible
in Figure~\ref{fig:graphs} that neither pairwise metric alone can recover.
A single combined adjacency---e.g.\ a weighted average of the three score
matrices---would merge complementary signals before the attention heads can
differentiate them, discarding the structural information that enables specialisation.
The pairwise union of any two graphs covers strictly more sensor pairs than either
alone, and the three-graph union covers $32.8\%$ of directed pairs on PEMS04
while pruning the remaining $67.2\%$---a coverage that no single-type graph achieves.
Table~\ref{tab:ablation} validates this empirically: the 3-relation composition
strictly outperforms every single-relation variant on both PEMS04 and PEMS08.

\textbf{Cyclic head assignment across all layers.}
Assigning head $i$ to relation $r_i{=}i\bmod 3$ ensures that every encoder layer
simultaneously uses all three relation types rather than dedicating individual layers
to individual relations.
This prevents a representation bottleneck in which a structural signal is available
only in early or late layers, forcing the model to integrate all three dependency
types at each level of the encoding hierarchy.
Empirically, this mirrors the iTransformer finding that layer normalisation applied
per-variate rather than per-timestep avoids cross-variate smearing at each layer:
here, cycling masks per-head rather than per-layer avoids relation-type smearing
across the hierarchy.
Since the mask array is broadcast across all layers, the cyclic assignment adds no
per-layer memory or parameter overhead.
With $h{=}4$ heads (canonical protocol), partial-correlation receives two dedicated
heads (heads 0 and 3) and Pearson and lead-lag each receive one; with $h{=}8$
heads each relation receives 2--3 heads.
The sensitivity analysis (Table~\ref{tab:ksens}) confirms that the method is robust
to the head allocation implied by different $h$ values.

%-----------------------------------------------------------------------

\subsection{Further Analysis}
\label{sec:stability}

\textbf{Seed stability.} Across all canonical seed runs in our released
per-seed data, neither \itrans{} nor \mr{} diverges, and both are
seed-stable: worst-case relative seed spread is $7\%$ for \itrans{} and
$3\%$ for \mr{}. The contrast is with \emph{learned} cross-channel
attention: CrossFormer's seed spread reaches $44\%$ at PEMS07 pred-12 and
$60\%$ at PEMS04 pred-48 under the identical protocol (seeds
$0.129$--$0.207$), making single-seed CrossFormer numbers unreliable
exactly where its means look strongest. Structural masking thus does not
merely match the backbone's stability---it avoids the instability that
comes with learning the channel graph instead of precomputing it.

A mechanism consistent with this: at initialisation each full-attention
weight is $\alpha_{ij}{\approx}1/N$ and off-diagonal softmax Jacobian
entries are $O(1/N^2)$; at $N{=}883$ masked attention ($k{=}20$) enjoys a
$(N/k)^2{\approx}1949\times$ larger gradient scale, and the fixed mask caps
the maximum logit gap, precluding the runaway sharpening available to a
learned router.

Table~\ref{tab:cost} shows that the mask mechanism is effectively free.
\mr{} has \emph{identical} parameter count to \itrans{} (the masks are
non-learnable buffers) and incurs only
1--3\% additional wall-clock time from the mask broadcast and fused masked softmax.
Graph precomputation is a one-time offline step (under 90\,s for PEMS07) that does
not affect inference latency; the mask array occupies $\approx2.3$\,MB at $N{=}883$,
negligible relative to the 40\,MB model.


\textbf{Scope and applicability.}
\label{sec:noharm}
\mr{} benefits datasets satisfying three conditions: $N$ must be large enough for
noise reduction from selective attention to be material; channels must exhibit genuine
statistical co-movement so the precomputed graphs are non-trivially populated; and
forecast targets at longer horizons must depend on spatial propagation that
short-range autocorrelation cannot capture alone.
All four PEMS benchmarks satisfy these criteria, and the method is applicable in
principle to other structured domains—electricity load networks (buildings sharing
distribution transformers), cross-asset financial series (equity factor structure), and
spatially extended atmospheric dynamics in climate forecasting—where training-set
statistics yield non-trivially populated relation graphs without requiring physical topology.
For small-$N$ datasets ($N\lesssim 50$) the benefit diminishes toward noise
(Sec.~\ref{sec:xdomain}). The sharper boundary is domains with diffuse,
non-stationary channel dependencies: on Traffic ($N{=}862$, Bay-Area-wide loop
detectors with non-local correlations), \mr{} is $9$--$16\%$ \emph{worse} than
\itrans{} across horizons 96--720 --- when the estimated communities do not
correspond to a physically grounded dependency structure, the fixed prior
removes capacity rather than noise. We report this negative result to
delimit the method's scope: precomputed relational masks are a prior for
\emph{structured} sensor networks, not a universal add-on.

\textbf{Implementation and reproducibility.}
\label{sec:repro}
Table~\ref{tab:hyperparams} lists shared hyperparameters; \mr{} and \itrans{} are
identical in every respect except for the mask buffer, enabling an exact like-for-like
comparison.


Code, precomputed masks, and per-seed raw results (JSON) will be released upon acceptance; an anonymized archive is provided as supplementary material.

%-----------------------------------------------------------------------
\textbf{Cross-domain experiments.}
\label{sec:xdomain}
To assess whether the gains on PEMS traffic benchmarks extend to non-traffic domains,
we evaluate \mr{} against \itrans{} on Solar-Energy ($N{=}137$ photovoltaic panel
sensors, Alabama), ETTh1, and ETTm1 ($N{=}7$, hourly and 15-minute oil-transformer
readings), using $k{=}20$ for Solar and $k{=}3$ for ETT; all other hyperparameters
follow the standard protocol (Table~\ref{tab:hyperparams}).


On Solar-Energy, despite $N{=}137$ and a union density (${\approx}14\%$) comparable
to PEMS07, pred-96 yields a slight \emph{degradation} ($-2.5\%$).
Unlike road sensors, photovoltaic panels share a single dominant signal (solar
irradiance) that renders the three graph types nearly redundant rather than
complementary; when all three masks converge to similar structure, the relational
heads lose their division of labour and offer no benefit over full attention.
On ETTh1 and ETTm1 ($N{=}7$, $k{=}3$), the mask union density is high
(${\approx}82\%$), leaving little room for sparsification; both datasets yield
gains within noise (ETTh1: ${\approx}0\%$; ETTm1: $+0.4\%$; ETTh2: ${\approx}0\%$).

Taken together, the cross-domain results confirm that \mr{}'s benefit requires
\emph{structurally diverse} multi-relational dependencies where the three graph types
are genuinely complementary---a condition met by traffic sensor networks (where direct
adjacency, marginal co-movement, and community dynamics are three distinct phenomena)
but not by networks with a single dominant shared driver
(Solar-Energy, where irradiance drives all sensors uniformly).

%-----------------------------------------------------------------------
\textbf{Extended benchmark.}
\label{sec:extended}
We further evaluate on ECL ($N{=}321$ electricity-consumption buildings, $k{=}20$),
where 321 consumers with distinct residential, commercial, and industrial schedules
produce complementary graph types.
\mr{} achieves $+$2.1\% improvement at pred-96 ($0.1477\to0.1446$, mean 3 seeds);
over the full horizon sweep the effect is neutral ($+1.4\%$ at 192, $-2.4\%$ at
336, $-0.7\%$ at 720; 4-horizon mean essentially tied), indicating that ECL's
structure is exploitable at short horizons but not stationary enough to help
beyond them. S-Mamba~\cite{wang2024smamba} reports a published ECL average of
0.170 as a reference point.
Weather ($N{=}21$) and ETTh1/ETTm1/ETTh2 ($N{=}7$) show gains within noise
(${\approx}0\%$), consistent with the structural diversity hypothesis.

\input{draft_sections/scope}
