% =====================================================================
% draft_sections/framework_intro.tex
% Reframe of the Introduction contribution: MRMA as a BACKBONE-AGNOSTIC
% FRAMEWORK, not a single model. Drop-in for the current Introduction's
% contribution paragraph + bulleted list (main.tex lines ~198-233).
%
% Suggested preamble macros for the integrating agent (add near \mr/\itrans
% at main.tex line ~59). If not added, replace the macro uses below with the
% \textsc{...} spellouts shown in the comments.
%   \newcommand{\mrma}{\textsc{MRMA}}                 % the mechanism
%   \newcommand{\crossf}{\textsc{CrossFormer}}        % backbone 2
%   \newcommand{\secf}{\textsc{SECrossformer}}       % instantiation 2
%   \newcommand{\structmamba}{\textsc{StructMamba}}   % instantiation 3
% (\mr and \itrans already exist.)
% =====================================================================

The lever we contribute is not a model but a \emph{mechanism}.
We propose \emph{multi-relational masked attention} (MRMA): three relation
graphs---partial-correlation, Pearson correlation, and an MDL-guided lead-lag
community partition obtained by approximately minimising 2D structural
entropy~\cite{li2016structural}---are precomputed once from the training split
and applied as per-head \emph{binary attention masks}.
Because the graphs are fixed buffers rather than learned tensors, MRMA adds
\emph{zero parameters} and touches only the pre-softmax logits of an existing
channel-mixing attention: it is a drop-in that leaves the surrounding backbone
otherwise unchanged.
This makes the contribution \emph{backbone-agnostic}---the same three graphs
condition any architecture whose forward pass mixes channels through an
attention (or attention-like gating) matrix, without redesigning the backbone.

We substantiate this claim with three independent instantiations spanning three
distinct architecture families, each compared against its own unmodified
backbone under an identical protocol.
(i)~\mr{}${=}$\itrans{}${+}$MRMA masks \itrans{}'s cross-channel self-attention
and is our primary instantiation.
(ii)~\textsc{SECrossformer}${=}$\textsc{CrossFormer}~\cite{zhang2023crossformer}${+}$MRMA
masks CrossFormer's learned cross-dimension attention.
(iii)~\textsc{StructMamba}${=}$a GRU-based state-space forecaster${+}$MRMA
carries the same graphs into a recurrent state-space backbone whose no-mask
variant is the fair baseline.
Across all four PEMS traffic benchmarks (PEMS03/04/07/08, $N{=}170$ to $883$),
each instantiation improves on its own backbone---\mr{} in all 16
dataset--horizon cells (mean $+10.9\%$, up to $+21.5\%$ at PEMS04 pred-96),
\textsc{SECrossformer} in 12 of 16 (mean $+17.0\%$), and \textsc{StructMamba}
in 14 of 16 (mean $+4.4\%$)---and together the three instantiations hold the
best (lowest-MSE) result in 12 of 16 cells against ten baselines.
The improvement is therefore a property of the \emph{mask mechanism}, reproduced
across backbones, rather than a property of any single tuned architecture.

Crucially, the gains come from the \emph{structure} of the prior, not from
sparsity per se: across all 16 MultiRel ablation cells, the unstructured
all-ones (fully connected) mask is the weakest variant---removing the structural
content while retaining the masking machinery erases the benefit.
We are equally explicit about where the framework does \emph{not} help.
The masks improve forecasting on flow-type traffic sensor networks, where the
precomputed graphs recover a genuine local dependency structure, but they reduce
to the backbone on diffuse or low-channel data (electricity, exchange-rate and
NASDAQ finance) and on traffic-\emph{speed} data, where no such structure is
recovered.
We report this scope boundary as a first-class finding
(Section~\ref{sec:scope}), not a footnote.

\medskip
\noindent\textbf{Contributions.}
\begin{itemize}
  \item \textbf{A backbone-agnostic mechanism.} We introduce MRMA, a
        zero-parameter, drop-in channel-mixing mask built from three
        complementary precomputed relation graphs (partial-correlation, Pearson,
        and an MDL-grounded 2D-structural-entropy lead-lag partition), and
        formalise it independently of any host architecture
        (Section~\ref{sec:method}).
  \item \textbf{Three instantiations, three architecture families.} We
        instantiate MRMA in an inverted Transformer (\mr{}${=}$\itrans{}${+}$MRMA),
        a cross-dimension Transformer
        (\textsc{SECrossformer}${=}$\textsc{CrossFormer}${+}$MRMA), and a
        GRU-based state-space forecaster
        (\textsc{StructMamba}${=}$GRU-SSM${+}$MRMA), demonstrating that the same
        graphs transfer unchanged (Section~\ref{sec:instantiations}).
  \item \textbf{Consistent backbone improvement.} Each instantiation beats its
        own unmodified backbone on PEMS (\mr{} 16/16, mean $+10.9\%$;
        \textsc{SECrossformer} 12/16, mean $+17.0\%$; \textsc{StructMamba}
        14/16, mean $+4.4\%$), and the three together attain the best cell in
        12 of 16 against ten baselines (Section~\ref{sec:frameworkresults}).
  \item \textbf{Structure, not sparsity---with an honest scope.} Ablations show
        the all-ones mask is the weakest variant in all 16 MultiRel cells,
        isolating the structural prior as the source of the gain; and we
        delimit the framework's applicability to structured flow-type sensor
        networks, reporting the diffuse/low-channel/speed regimes where it
        reduces to the backbone (Section~\ref{sec:scope}).
\end{itemize}
